Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-18 paper.tex
Draft of explaining example

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts, bbm}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{placeins}
\usepackage[hidelinks]{hyperref}

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify %~\citep{wong_general_2023,kelleher_efficient_2018} 
the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on it;
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, Fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG inference methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
% JK I think we can but this out without-loss now given Halley's paper?
% Currently,
% \argweaver{} \citep{rasmussen_genome-wide_2014},
% \tsinfer{} \citep{kelleher_inferring_2019},
% \kwarg{} \citep{ignatieva_kwarg_2021}, and
% \texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
% including unary portions of coalescence nodes.
% However, it is unclear to what degree these are well-inferred,
% as most evaluation of inferred ARGs thus far has focused on marginal trees
% (e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
% but see \citet{deng2021distribution})
% and thus completely ignore such unary nodes
% (and their implications for inherited haplotypes).
See~\cite{fritze2024forest} for more discussion on unary nodes
and haplotype aware ARG comparison metrics.

@@

Our second example shows a different aspect of this robustness to missing 
data. In Fig~\ref{fig:lik-surface} we can see the likelihood-ratio curves
obtained when we compute the likelihood of an example simulation
at varying degrees of completeness.[JK: WIP]

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/likratio.png}
    \caption{Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

In line with the observation that the presence of locally unary nodes 
(see section \ref{par:recording}) is crucial for the equivalence between
the graph or local trees view of an ARG, one can gradually simplify down
a simulated ARG by gradually removing the various types of locally unary nodes.
Figure \ref{fig:lik-surface} shows the impact of these simplification steps
on the likelihood for a single simulated ARG.
The log likelihood ratio is shown to improve comparability.

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts, bbm}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{placeins}
% \usepackage[hidelinks]{hyperref}

@@

These locally unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify 
the ARG by removing all recombinant nodes
($A$, $B$, and $C$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombinant nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombinant nodes we no longer know the precise time of each recombination event,
but we do still have constraints on it;
for instance, we know that $C$ happened somewhere between node $f$ and node $h$.
More generally, a recombinant node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, Fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG inference methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
% JK I think we can but this out without-loss now given Halley's paper?
% Currently,
% \argweaver{} \citep{rasmussen_genome-wide_2014},
% \tsinfer{} \citep{kelleher_inferring_2019},
% \kwarg{} \citep{ignatieva_kwarg_2021}, and
% \texttt{SINGER} \citep{deng_robust_2024} estimate ARGs 
% including unary portions of coalescence nodes.
% However, it is unclear to what degree these are well-inferred,
% as most evaluation of inferred ARGs thus far has focused on marginal trees
% (e.g., \citet{brandt2022evaluation, kelleher_inferring_2019},
% but see \citet{deng2021distribution})
% and thus completely ignore such unary nodes
% (and their implications for inherited haplotypes).
See~\cite{fritze2024forest} for more discussion on unary nodes
and haplotype aware ARG comparison metrics.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/likratio.png}
    \caption{TODO add A/B and change ``unary regions''.
    Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

Our second example shows a different aspect of this robustness to missing 
data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
obtained when we compute the likelihood of an example simulation,
in which the ARG is at three different levels of completeness
(illustrated in Fig~\ref{fig:lik-surface}B).
We first run a simulation under the SMC
using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
ARG explicitly identifies all of the recombination and common 
ancestor events that occurred in the simulation, and 
we can therefore compute a likelihood under the HKY formulation
(for example, using the \texttt{msprime.log\_arg\_likelihood} 
function). The example full ARG in 
Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
unary''~\citep{wong_general_2023}. Nodes 6 and 7 correspond to the 
left and right parents of the recombination event causing the 
breakpoint at position 536~\cite[we need two nodes to uniquely identify
 the passage of ancestral material through recombination 
events; see][]{baumdicker_efficient_2021}, and therefore present
in only those trees. 
Node 8 is a coalescent node in the right-hand tree, but unary
in the left-hand tree, marking the passage of the lineage through
this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
and compute the likelihood using equation~\eqref{eq:full-lik},
there is very little difference in the likelihood ratio
curve from the exact value with complete input. 
Most importantly, the minimum occurs at the true value of the 
recombination parameter.
When we further simplify
the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
we obtain an ARG in the standard format used by \texttt{msprime}.
Although there is substantial information about the correlation
structure of trees in this representation, we lose important
information about recombination. This is demonstrated in 
Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
is clearly biased.
